\pdftitle

### 2018-05-13 maxNt_v2.tex
V2

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-13T13:23:37+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\def\arxivp{}
\def\mparcp{}
\def\philscip{}
\def\biorxivp{}
\newcommand*{\arxivsi}{\texttt{arXiv} eprints available at \url{http://arxiv.org/}.\\}
\newcommand*{\mparcsi}{\texttt{mp\_arc} eprints available at \url{http://www.ma.utexas.edu/mp_arc/}.\\}
\newcommand*{\philscisi}{\texttt{philsci} eprints available at \url{http://philsci-archive.pitt.edu/}.\\}
\newcommand*{\biorxivsi}{\texttt{bioRxiv} eprints available at \url{http://biorxiv.org/}.\\}
\newcommand*{\arxiveprint}[1]{\global\def\arxivp{\arxivsi}%\citeauthor{0arxivcite}\addtocategory{ifarchcit}{0arxivcite}%eprint
\texttt{\urlalt{http://arxiv.org/abs/#1}{arXiv:\hspace{0pt}#1}}%
%\texttt{\href{http://arxiv.org/abs/#1}{\protect\url{arXiv:#1}}}%
%\renewcommand{\arxivnote}{\texttt{arXiv} eprints available at \url{http://arxiv.org/}.}
}
\newcommand*{\mparceprint}[1]{\global\def\mparcp{\mparcsi}%\citeauthor{0mparccite}\addtocategory{ifarchcit}{0mparccite}%eprint
\texttt{\urlalt{http://www.ma.utexas.edu/mp_arc-bin/mpa?yn=#1}{mp\_arc:\hspace{0pt}#1}}%
%\texttt{\href{http://www.ma.utexas.edu/mp_arc-bin/mpa?yn=#1}{\protect\url{mp_arc:#1}}}%
%\providecommand{\mparcnote}{\texttt{mp_arc} eprints available at \url{http://www.ma.utexas.edu/mp_arc/}.}
}
\newcommand*{\philscieprint}[1]{\global\def\philscip{\philscisi}%\citeauthor{0philscicite}\addtocategory{ifarchcit}{0philscicite}%eprint
\texttt{\urlalt{http://philsci-archive.pitt.edu/archive/#1}{PhilSci:\hspace{0pt}#1}}%
%\texttt{\href{http://philsci-archive.pitt.edu/archive/#1}{\protect\url{PhilSci:#1}}}%
%\providecommand{\mparcnote}{\texttt{philsci} eprints available at \url{http://philsci-archive.pitt.edu/}.}
}
\newcommand*{\biorxiveprint}[1]{\global\def\biorxivp{\biorxivsi}%\citeauthor{0arxivcite}\addtocategory{ifarchcit}{0arxivcite}%eprint
\texttt{\urlalt{http://biorxiv.org/content/early/#1}{bioRxiv:\hspace{0pt}#1}}%
%\texttt{\href{http://arxiv.org/abs/#1}{\protect\url{arXiv:#1}}}%
%\renewcommand{\arxivnote}{\texttt{arXiv} eprints available at \url{http://arxiv.org/}.}
}

@@

\author{
  P.G.L. Porta Mana\\\texttt{piero.mana@ntnu.org}
  \And
  V. Rostami\\\texttt{v.rostami@fz-juelich.de}
  \And
E. Torre\\\texttt{torre@ibk.baug.ethz.ch}
  % David S.~Hippocampus\thanks{Use footnote for providing further
  %   information about author (webpage, alternative
  %   address)---\emph{not} for acknowledging funding agencies.} \\
  % Department of Computer Science\\
  % Cranberry-Lemon University\\
  % Pittsburgh, PA 15213 \\
  % \texttt{hippo@cs.cranberry-lemon.edu} \\
  %% examples of more authors
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

@@

Imagine that we have recorded the firing activity of a hundred neurons from
a particular brain area. This scenario is a concrete possibility thanks to
recent electrophysiological techniques \cite{berenyietal2014}. We bin and
binarize the recorded activity into sequences of $0$s (inactive) and $1$s
(firing) \cite{caianiello1961,caianiello1986}. We notice that groups of two
or more $1$s can appear in any time bin. We may ask several questions
looking at these groups. A recurring question is of this kind: can the
presence of groups of three or more simultaneous firings be
\enquote{explained} in terms of groups of just two or three? This kind of
question is very vague until we define what we mean with \enquote{explain}
or similar words.

A possible, precise definition is as follows. We want to predict, \ie\
assign a probability to, the population activity in a time bin, given the
population activities in the other bins:
\begin{equation}
  \label{eq:general_prediction}
  \p(\text{\small activity at one time bin} \|
  \text{\small activities at the remaining bins},\text{\small other info}).
\end{equation}
We can say that the simultaneous firing of at most two neurons
\enquote{explains} the activity if only the values of the first and second
empirical moments of the activities are relevant for the probability above:
\begin{multline}
  \label{eq:informal_sufficiency}
  \p(\text{\small activity at one time bin} \|
  \text{\small activities at the other bins},\text{\small other info}) ={}\\
  \p(\text{\small activity at some time bin} \|
  \text{\small 1st \amp\ 2nd empirical moments},\text{\small other info});
\end{multline}
in statistical terms, the first and second moments are \emph{sufficient
  statistics}. Other definitions are possible, of course; the definition
above has the advantage of being mathematically very precise. In fact it
determines, from just the three basic laws of the probability calculus, the
mathematical form of the probability for the activity at any number of bins:
\begin{equation}
  \label{eq:sufficiency_maxent_family}
  \begin{gathered}
  \p(\text{\small activity at bins $1,2,\dotsc$} \|
  \text{\small other info}) =
  \int\Bigl[\prod_i 
    \pf(\text{\small activity at bin } i \|\ytheta) \Bigr]\,
 \pf(\ytheta\| \text{\small other info})\,
  \di\ytheta,
  \\[\jot]
  \text{with \enquote{$\pf(\text{\small activity at time bin } i \|\ytheta)$}
    having maximum-entropy form with multipliers $\ytheta$};
\end{gathered}
\end{equation}
The proof that the assumption of sufficient statistics leads
to the probability expression above is the Koopman-Pitman theorem
\cite{koopman1936,pitman1936,darmois1935} and can be found in
\textcite[\sect~4.5]{bernardoetal1994_r2000} and references therein. It can
be extended to higher moments and other statistics.

When many data are available (and unfortunately also when they aren't), the
conditional density for the Lagrange multipliers,
$\pf(\ytheta\| \text{data, other info})$ is strongly peaked at the unique
values determined by the inverse problem \cite{meadetal1984}, and the
probability for the population activity is well-approximated by a
maximum-entropy distribution \cite{jaynes1957,sivia1996_r2006,meadetal1984}
with those Lagrange multipliers. This kind of distributions has been used
in neuroscience at least since the 1990s to study correlations from diverse
points of view, for diverse purposes, and with data recorded from diverse
brain areas
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015}.
Part of this literature focused on the range of applicability of maximum-entropy
distributions
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013}.

This kind of studies would have little significance if we did not expect
their results to generalize beyond the particular recorded sample. In other
words, we implicitly or explicitly expect that the conclusions drawn with
maximum-entropy methods about the recorded neurons should somehow
generalize or extrapolate at least to whole areas those neurons are recorded from,
\eg\ retina or motor cortex. We are thus assuming that the recorded sample
is a \emph{representative sample} of that area.

Whether such extrapolation makes biological sense is a difficult question,
to be analysed case by case. The purpose

The point that we want to stress is that

From the observation of these samples we expect
to learn something about all other neurons in the same brain area. That is,
we assume that they are a \enquote{representative sample}.

This note is mainly addressed to neuroscientists interested in \me\
methods, but we would be pleased if its discussion of the probability of
\enquote{sampling} were useful to neuroscientists that use other
statistical methods to study the physical and dynamical characteristics of
brain areas via neuronal recordings.

Recent electrophysiological techniques \cite{berenyietal2014} in fact allow
experimenters to record samples comprising even a couple hundreds neurons
from specific brain areas. From the observation of these samples we expect
to learn something about all other neurons in the same brain area. That is,
we assume that they are a \enquote{representative sample}. We do not
elaborate on the various important purposes of such recordings here, but
stress that these sample sizes can be considered \enquote{large} because
their statistical analysis requires considerable computational power.

These computational costs are one, probably not the earliest, of the many
reasons why \me\ methods have been introduced in neuroscience. It would be
useful to somehow compress the statistical wealth of large neuronal
recordings into few quantities, like sample moments for example. This
compression might also entail interesting biological or functional
properties of neuronal activity. The standard \me\ method
\cite{jaynes1957,sivia1996_r2006,meadetal1984} accomplishes this kind of
compression: it associates a unique probability distribution with few
experimental quantities. But this is only one of its uses. It is also used
for various information-theoretic purposes or to generate reference
probability distributions
\cite{bohteetal2000,schneidmanetal2006,shlensetal2006,roudietal2009b,mackeetal2011,tkaciketal2014b,shimazakietal2015}.
In all these uses the \me\ distribution is chosen as the \enquote{maximally
  noncommittal} one \cite{jaynes1963}. This adjective means little without
further technical characterizations. Different works give different
characterizations, but in this paper we will use the quoted expression as
an umbrella term for all of them. Our results will not depend on the
specific characterization of \enquote{noncommittal}.

@@

In this note we show that we must face a dilemma if we want to apply the
\me\ method to a representative sample to find a maximally noncommittal
distribution.

The dilemma is this. We can apply the \me\ method to the sample, using a
specified set of experimental constraints, and generate a probability
distribution for the state of the sample. But our sample is representative
of a larger population, in which it is biologically and functionally
embedded. We can apply the \me\ method to the larger population, using the
same constraints, and generate a probability distribution for its larger
state, and then find the distribution for the sample by marginalization.
Either application seems to have some commendable features. However,
\emph{the distributions obtained from these two applications differ}. It
goes without saying that if one is \enquote{noncommittal}, the other must
be \enquote{committal}. Which choice is most meaningful?

@@

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}
The conditional probability in the last formula is a hypergeometric
distribution $\ypp(\yrs\|\yRf)$, typical of \enquote{drawing without
  replacement} problems. The combinatorial proof of the formulae above is
in fact the same as for this class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
Our initial symmetric knowledge should intuitively also apply to the
sample; indeed, the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

@@

The conditional probability $\ypp(\yrs,\yRf)$ relates the spaces of the
sample average $\yXf\in\set{0,\dotsc,N}$ and of the population average
$\yxs\in\set{0,\dotsc,n}$ in a special way. It is a coarsening projector of
any probability $p$ for $\yXf$ onto a marginal probability $p_*$ for
$\yxs$:
\begin{equation}
  \label{eq:projection_P}
  % \ypp\colon p(\yXf=\dotv) \to
  p_*(\yxs=\yrs)
  = \sum_{N\yRf=0}^N \ypp(\yrs\|\yRf) p(\yXf=\yRf).
\end{equation}
Conversely it also a pulls back expectations of functions $f$ of the sample
average $\yxs$ to expectations of functions $f^*$ of the population average
$\yXf$:
\begin{equation}
  \label{eq:pullback_P}
%  \ypp^*\colon f(\yxs=\dotv) \to f^*(\yXf=\dotv) \defd \sum_{N\yrs=0}^n
%  f(\yrs)\ypp(\yrs\|\dotv),
  \begin{gathered}
    f^*(\yXf) \defd \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yXf),
    \\
   \expeb{f(\yxs)}= \expeb{f^*(\yXf)}
    =
    \sum_{n\yrs=0}^n f(\yrs)\,\p(\yxs=\yrs \| \yH)
    =
    \sum_{N\yRf=0}^N f^*(\yRf)\,\p(\yXf=\yRf \| \yH).
  \end{gathered}
\end{equation}

\begin{figure}[!t]
\centering
\includegraphics[width=0.9\linewidth]{pop_sample_projection3.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
\ypp(\yrs\|\yRf) \defd  \raisebox{0pt}[0pt][1ex]{}\binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}$ for $N=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yRf)$, see
\fig~\ref{fig:hypergeom_proj}, reveals that it is a sort of \enquote{fuzzy
  identity matrix} between the $\yXf$-space $\set{0,\dotsc,N}$ and
$\yxs$-space $\set{0,\dotsc,n}$. When $n=N$ it is the identity matrix. We
thus have that
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \approx \p(\yXf=a),\qquad
\expeb{f(\yxs)} \approx \expeb{f(\yXf)}.
\end{equation}
These are only very approximate equalities: they may miss important
features of the two probability distributions. In the next section we will
in fact emphasize their differences. If the distribution for the population
average $\yXf$ is bimodal, for example, the bimodality can be lost in the
distribution for the sample average $\yxs$, owing to the coarsening effect
of $\ypp(\yrs\|\yRf)$.

@@

% [3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of
% learning and recall at excitatory recurrent synapses and cholinergic
% modulation in rat hippocampal region CA3. {\it Journal of
%   Neuroscience} {\bf 15}(7):5249-5262.
\defbibnote{postnote}{\small\par\medskip\noindent{\footnotesize% Note:
\arxivp \mparcp \philscip \biorxivp}%
}

@@

\printbibliography[postnote=postnote]

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-13T22:11:43+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\newcommand*{\arxiveprint}[1]{%\global\def\arxivp{\arxivsi}%\citeauthor{0arxivcite}\addtocategory{ifarchcit}{0arxivcite}%eprint
\texttt{\urlalt{https://arxiv.org/abs/#1}{arXiv:\hspace{0pt}#1}}%
%\texttt{\href{http://arxiv.org/abs/#1}{\protect\url{arXiv:#1}}}%
%\renewcommand{\arxivnote}{\texttt{arXiv} eprints available at \url{http://arxiv.org/}.}
}
\newcommand*{\mparceprint}[1]{%\global\def\mparcp{\mparcsi}%\citeauthor{0mparccite}\addtocategory{ifarchcit}{0mparccite}%eprint
\texttt{\urlalt{http://www.ma.utexas.edu/mp_arc-bin/mpa?yn=#1}{mp\_arc:\hspace{0pt}#1}}%
%\texttt{\href{http://www.ma.utexas.edu/mp_arc-bin/mpa?yn=#1}{\protect\url{mp_arc:#1}}}%
%\providecommand{\mparcnote}{\texttt{mp_arc} eprints available at \url{http://www.ma.utexas.edu/mp_arc/}.}
}
\newcommand*{\philscieprint}[1]{%\global\def\philscip{\philscisi}%\citeauthor{0philscicite}\addtocategory{ifarchcit}{0philscicite}%eprint
\texttt{\urlalt{http://philsci-archive.pitt.edu/archive/#1}{PhilSci:\hspace{0pt}#1}}%
%\texttt{\href{http://philsci-archive.pitt.edu/archive/#1}{\protect\url{PhilSci:#1}}}%
%\providecommand{\mparcnote}{\texttt{philsci} eprints available at \url{http://philsci-archive.pitt.edu/}.}
}
\newcommand*{\biorxiveprint}[1]{%\global\def\biorxivp{\biorxivsi}%\citeauthor{0arxivcite}\addtocategory{ifarchcit}{0arxivcite}%eprint
\texttt{\urlalt{http://biorxiv.org/content/early/#1}{bioRxiv:\hspace{0pt}#1}}%
%\texttt{\href{http://arxiv.org/abs/#1}{\protect\url{arXiv:#1}}}%
%\renewcommand{\arxivnote}{\texttt{arXiv} eprints available at \url{http://arxiv.org/}.}
}
\newcommand*{\osfeprint}[1]{%
\texttt{\urlalt{https://doi.org/10.17605/osf.io/#1}{doi:10.17605/osf.io/#1}}%
}

@@

\author{
  P.G.L. Porta Mana\\\texttt{piero.mana@ntnu.no}
  \And
  V. Rostami\\\texttt{v.rostami@fz-juelich.de}
  \And
E. Torre\\\texttt{torre@ibk.baug.ethz.ch}
  % David S.~Hippocampus\thanks{Use footnote for providing further
  %   information about author (webpage, alternative
  %   address)---\emph{not} for acknowledging funding agencies.} \\
  % Department of Computer Science\\
  % Cranberry-Lemon University\\
  % Pittsburgh, PA 15213 \\
  % \texttt{hippo@cs.cranberry-lemon.edu} \\
  %% examples of more authors
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

@@

Imagine that we have recorded the firing activity of a hundred neurons from
a particular brain area. This scenario is a concrete possibility thanks to
recent electrophysiological techniques \cite{berenyietal2014}. In 1985,
Gerstein, Perkel, \amp\ Dayhoff wrote:
\begin{quote}
  the very profusion of data produced in such experiments creates serious
  problems of analysis and interpretation; these problems are conceptual as
  well as practical. The principal conceptual problems are (1) defining
  cooperativity or functional grouping among neurons and (2) formulating
  quantitative criteria for recognizing and characterizing such
  cooperativity. \cite{gersteinetal1985}
\end{quote}
 More than thirty years later these problems are
still under investigation.

Some investigations in the literature proceed as follows. We bin in time
and binarize in intensity \cite{caianiello1961,caianiello1986} the recorded
activity, transforming it into a sequence of $0$s (inactive) and $1$s
(firing) for each neuron. Groups of two or more $1$s can appear in any time
bin. Then we ask a question of this kind: can the presence of groups of
three or more simultaneous firings be \enquote{explained} in terms of
groups of just two, or three, or\ldots? This kind of question is very vague
until we define what we mean with \enquote{explain} or similar words.
Usually such \enquote{explanation} has a probabilistic meaning: we wonder
whether the sole knowledge about two, or three, or more simultaneous
firings can yield as good predictions about the whole activity as a more
complete knowledge would do. To make this question concrete we must choose
a probabilistic model.

Maximum-entropy models
\cite{jaynes1957,jaynes1963,sivia1996_r2006,meadetal1984} have been used in
neuroscience with this kind of purpose and for diverse brain areas at least
since the 1990s
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015}.
Part of this literature focused on the range of applicability of \me\
distributions
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013,rostamietal2016_r2017}.
We assume that our readers are at least vaguely familiar with such models.
In brief, they build a probability distribution for neuronal activity using
only data about simultaneous firings of two neurons, or of three, and so
on. Maximum-entropy probability distributions can be introduced and justified in
three main ways:
\begin{enumerate}
\item from the maximum-entropy method
  \cite{jaynes1957,jaynes1963,sivia1996_r2006};
\item as asymptotic distributions of Bayesian models called
  \enquote{exchangeable} with entropic priors
  \cites{jaynes1986d_r1996,skilling1989b,skilling1990,rodriguez1989,rodriguez2002,portamana2017}[see
  also][\chap~VI]{schroedinger1946_r1948}%[pp.~52--55]
{jaynes1979b};
\item as asymptotic distributions of Bayesian models based on sufficient
  statistics
  \cite{jaynes1982,%[\sect~I]
jaynes1982b,portamana2017}.
\end{enumerate}

*** usually \enquote{explain} has a probabilistic meaning. We want to see
whether information about moments give reasonable predictions about the
rest. Necessity of making this precise. Three ways:

-- \me\ method,

-- Bayesian with entropic prior,

-- Bayesian sufficiency.

Applied to sample, all three are asymptotically equivalent. However, sample
is representative of population. Need to take full population into
consideration. This leads to different formulae, and also to
non-equivalence of third way. Here we focus on first two.

This kind of
question is very vague until we define what we mean with \enquote{explain}
or similar words.

A possible, precise definition is as follows. We want to predict, \ie\
assign a probability to, the population activity in a time bin, given the
population activities in the other bins:
\begin{equation}
  \label{eq:general_prediction}
  \p(\text{\small activity at one time bin} \|
  \text{\small activities at the remaining bins},\text{\small other info}).
\end{equation}
We can say that the simultaneous firing of at most two neurons
\enquote{explains} the activity if only the values of the first and second
empirical moments of the activities are relevant for the probability above:
\begin{multline}
  \label{eq:informal_sufficiency}
  \p(\text{\small activity at one time bin} \|
  \text{\small activities at the other bins},\text{\small other info}) ={}\\
  \p(\text{\small activity at some time bin} \|
  \text{\small 1st \amp\ 2nd empirical moments},\text{\small other info});
\end{multline}
in statistical terms, the first and second moments are \emph{sufficient
  statistics}. Other definitions are possible, of course; the definition
above has the advantage of being mathematically very precise. In fact it
determines, from just the three basic laws of the probability calculus, the
mathematical form of the probability for the activity at any number of bins:
\begin{equation}
  \label{eq:sufficiency_maxent_family}
  \begin{gathered}
  \p(\text{\small activity at bins $1,2,\dotsc$} \|
  \text{\small other info}) =
  \int\Bigl[\prod_i 
    \pf(\text{\small activity at bin } i \|\ytheta) \Bigr]\,
 \pf(\ytheta\| \text{\small other info})\,
  \di\ytheta,
  \\[\jot]
  \text{with \enquote{$\pf(\text{\small activity at time bin } i \|\ytheta)$}
    having \me\ form with multipliers $\ytheta$};
\end{gathered}
\end{equation}
The proof that the assumption of sufficient statistics leads
to the probability expression above is the Koopman-Pitman theorem
\cite{koopman1936,pitman1936,darmois1935} and can be found in
\textcite[\sect~4.5]{bernardoetal1994_r2000} and references therein. It can
be extended to higher moments and other statistics.

When many data are available (and unfortunately also when they aren't), the
conditional density for the Lagrange multipliers,
$\pf(\ytheta\| \text{data, other info})$ is strongly peaked at the unique
values determined by the inverse problem \cite{meadetal1984}, and the
probability for the population activity is well-approximated by a
\me\ distribution
\cite{jaynes1957,jaynes1963,sivia1996_r2006,meadetal1984} with those
Lagrange multipliers. This kind of distributions has been used in
neuroscience at least since the 1990s to study correlations from diverse
points of view, for diverse purposes, and with data recorded from diverse
brain areas
\cite{mackay1991,martignonetal1995,bohteetal2000,amarietal2003,schneidmanetal2006,shlensetal2006,mackeetal2009b,roudietal2009c,tkaciketal2009,gerwinnetal2009,mackeetal2011,mackeetal2011b,ganmoretal2011,granotatedgietal2013,tkaciketal2014b,moraetal2015,shimazakietal2015}.
Part of this literature focused on the range of applicability of
\me\ distributions
\cite{tkaciketal2006,roudietal2009,roudietal2009b,barreiroetal2010b,mackeetal2013}.

This kind of studies would have little significance if we did not expect
their results to generalize beyond the particular recorded sample. In other
words, we implicitly or explicitly expect that the conclusions drawn with
\me\ methods about the recorded neurons should somehow generalize or
extrapolate at least to whole areas those neurons are recorded from, \eg\
retina or motor cortex. We are thus assuming that the recorded sample is a
\emph{representative sample} of that area. Whether such extrapolation makes
biological sense is a difficult question, to be analysed case by case.

@@

But a problem remains even if we assume that the sample is representative
of its brain area. The problem lies in the way the \me\ methods
is typically used in the literature. In this note we want to discuss this
problem and offer a solution to it.

The problem is this. We can apply the \me\ method to the sample, choosing
some sufficient statistics, and generate a probability distribution for the
state of the sample. But we consider our sample to be representative of a
larger population, in which it is biologically and functionally embedded.
We can apply the \me\ method to the larger population, using the same kind
of sufficient statistics, and generate a probability distribution for its
larger state, and then find the distribution for the sample by
marginalization. Either application seems to have some commendable
features. However, \emph{the distributions obtained from these two
  applications differ}. It goes without saying that if one is
\enquote{noncommittal}, the other must be \enquote{committal}. Which choice
is most meaningful?

@@

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{where}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \shortintertext{with the conditional probability}
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}
The conditional probability in the last formula is a hypergeometric
distribution $\ypp(\yrs\|\yRf)$, typical of \enquote{drawing without
  replacement} problems. The combinatorial proof of the formulae above is
in fact the same as for this class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
Our initial symmetric knowledge should intuitively also apply to the
sample; indeed, the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

@@

The conditional probability
$\p(\yxs = \yrs \|\yXf=\yRf, \yH)\equiv\ypp(\yrs\|\yRf)$ relates the spaces
of the sample average $\yXf\in\set{0,\dotsc,N}$ and of the population
average $\yxs\in\set{0,\dotsc,n}$. It is a coarsening projector of any
probability $p$ for $\yXf$ onto a marginal probability, as
\eqn~\eqref{eq:subpop_average} shows. It also allows us to construct a
function of the population average $f^*(\yXf)$, given a function of the
sample average $f(\yxs)$:
\begin{equation}
  \label{eq:pullback_P}
  f^*(\yXf) \defd \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yXf),
\end{equation}  
with the property of having the same expectation value as the latter:
\begin{equation}
  \label{eq:same_expect}
  \expeb{f^*(\yXf)} \equiv
  \sum_{N\yRf=0}^N f^*(\yRf)\,\p(\yXf=\yRf \| \yH)
  =
  \sum_{n\yrs=0}^n f(\yrs)\,\p(\yxs=\yrs \| \yH)
  \equiv
  \expeb{f(\yxs)}. 
\end{equation}

\begin{figure}[!t]
\centering
\includegraphics[width=0.9\linewidth]{pop_sample_projection3.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
\ypp(\yrs\|\yRf) \defd  \raisebox{0pt}[0pt][1ex]{}\binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}$ for $N=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yRf)$,
\fig~\ref{fig:hypergeom_proj}, reveals that it is a sort of \enquote{fuzzy
  identity matrix} between the $\yXf$-space $\set{0,\dotsc,N}$ and
$\yxs$-space $\set{0,\dotsc,n}$. When $n=N$ it is the identity matrix. We
thus have that
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \approx \p(\yXf=a),\qquad
\expeb{f(\yxs)} \approx \expeb{f(\yXf)}.
\end{equation}
These are only very approximate equalities: they may miss important
features of the two probability distributions. In the next section we will
in fact emphasize their differences. If the distribution for the population
average $\yXf$ is bimodal, for example, the bimodality can be lost in the
distribution for the sample average $\yxs$, owing to the coarsening effect
of $\ypp(\yrs\|\yRf)$.

@@

% [3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of
% learning and recall at excitatory recurrent synapses and cholinergic
% modulation in rat hippocampal region CA3. {\it Journal of
%   Neuroscience} {\bf 15}(7):5249-5262.
% \defbibnote{postnote}{\small\par\medskip\noindent{\footnotesize% Note:
% \arxivp \mparcp \philscip \biorxivp}%
% }

@@

\printbibliography%[postnote=postnote]
